---
title: "EDA"
output:  
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyverse)
library(plotly)
```

## Clean data
```{r}
tornado_data <- read.csv("./data/1950-2023_actual_tornadoes.csv", na.strings = c("NA", "N/A", " "))

cleaned <- tornado_data |>
  filter(yr >= 2000) |>
  select(-tz, -stf, -ns, -sn, -sg, -f1, -f2, -f3, -f4) |>
  relocate(fc, .after = mag)|>
  drop_na()
  
# Optional
write.csv(cleaned, "cleaned_df.csv", row.names = FALSE)
```

```{r}
summary(cleaned)
```

- Most tornadoes have minimal impacts, with many records showing zero injuries, fatalities, or losses.
- A few extreme events contribute significantly to the maximum values in injuries, fatalities, and losses.
- Tornado path dimensions show high variability, with extreme widths and lengths in rare cases.

```{r}
skimr::skim(cleaned)  
```

#### **Magnitude (`mag`)**:
- Mean magnitude is 0.24, indicating that the majority of tornadoes are mild (EF0-EF2).
- The dataset includes a small percentage of extreme events (EF3-EF5), which likely contribute to severe damages.

#### **Path Dimensions (`len`, `wid`)**:
- Tornadoes typically have short paths, with most being less than 2 miles in length. However, a few extend up to 168.5 miles.
- Path widths show greater variability, with a median of 50 yards and rare cases exceeding 4,500 yards.

#### Concluding Observations
- **Low-impact majority**: Most tornadoes in the dataset result in no injuries, fatalities, or monetary losses.
- **Skewed distributions**: A small number of high-impact tornadoes dominate the dataset's extreme values.
- **Path variability**: Tornado paths show significant variability, with a few spanning extraordinary lengths and widths.
- **Data completeness**: The cleaned dataset is complete, with no missing values in any variable, enabling robust exploratory analysis.


### EDA

## Tornado Distribution

Annual trend of tornadoes from 2000 to 2023

```{r}
annual_trend = cleaned %>%
  group_by(yr) %>%
  summarise(count = n()) 

annual_plot = ggplot(annual_trend, aes(x = yr, y = count)) +
geom_line(color = "blue") +
geom_point()+
theme_minimal() +
labs(title = "Annual Number of Tornadoes (2000-2023)", x = "Year", y = "Count")

ggplotly(annual_plot)

```

Based on this plot, we found that there is no obvious trend according to years. There are fewer tornadoes around 2014 and there are higher tornadoes occurrence around 2004.

Monthly Distribution of Tornadoes

```{r}
monthly_distribution = cleaned %>%
  group_by(mo) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

print(monthly_distribution)

monthly_plot= ggplot(monthly_distribution, aes(x = factor(mo), y = count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Monthly Distribution of Tornadoes", x = "Month", y = "Count")
ggplotly(monthly_plot)
```

This monthly distribution of tornadoes is unimodal and slightly right-skewed, which concentrated within the range of April to June and has a peak in May.

Seasonal Distribution of Tornadoes

```{r}
seasonal_distribution = cleaned %>%
  mutate(
    season = case_when(
      mo %in% c(12, 1, 2) ~ "Winter",
      mo %in% c(3, 4, 5) ~ "Spring",
      mo %in% c(6, 7, 8) ~ "Summer",
      mo %in% c(9, 10, 11) ~ "Fall"
    ),
    season = factor(season, levels = c("Spring","Summer","Fall", "Winter"))
  )%>%
  group_by(season) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

print(seasonal_distribution)

seasonal_plot = seasonal_distribution %>%
  ggplot(aes(x = season, y = count, fill = season)) +
  geom_bar(stat = "identity") +
  labs(title = "Seasonal Tornado Counts", x = "Season", y = "Number of Tornadoes") +
  theme_minimal()

ggplotly(seasonal_plot)
```

From this histogram, Spring has more tornado occurrences which is more than 10000 tornadoes and winter is the least active season which is around 2900 tornadoes. The activity of tornado occurrences is decreasing seasonally.

What is the total number of tornadoes for each state?

```{r}
state_tornado_count_map <- cleaned |>
  group_by(st) |>
  summarize(total_tornadoes = n()) |>
  plot_ly(
    type = 'choropleth',
    locations = ~st,
    locationmode = 'USA-states',
    z = ~total_tornadoes,
    colorscale = "Viridis", 
    zmin = 0,               
    zmax = max(cleaned$st, na.rm = TRUE), 
    colorbar = list(title = "Total Tornadoes")
  ) |>
  layout(
    title = "Total Tornado Numbers by State",
    geo = list(scope = "usa")
  )

state_tornado_count_map
```

- **Texas (TX)** has the highest tornado count, exceeding **3,000**, reflecting its large size and frequent exposure to tornado-prone weather patterns.
- States in **Tornado Alley**, such as **Oklahoma (OK)**, **Kansas (KS)**, and **Nebraska (NE)**, also exhibit high tornado frequencies.
- **Eastern states**, including **Mississippi (MS)** and **Alabama (AL)**, show moderate tornado counts.
- **Western states**, such as **Nevada (NV)** and **California (CA)**, have the lowest tornado counts due to less favorable conditions for tornado formation.

## Tornado Impact

Relationship between tornado magnitude and its impact on injuries, fatalities and proporty loss.

```{r}
impact_plot = cleaned %>%
  pivot_longer(cols = c(inj, fat, loss), 
               names_to = "impact", 
               values_to = "value") %>%
  ggplot(aes(x = as.factor(mag), y = value, fill = impact)) +
  geom_boxplot(outlier.colour = "red", outlier.size = 0.8 ) +
  facet_wrap(~impact, 
             scales = "free", 
             labeller = labeller(impact = c(
    "inj" = "Injuries",
    "fat" = "Fatalities",
    "loss" = "Property Loss"
  ))) +
  labs(title = "Tornado Impacts by Magnitude", 
       x = "Tornado Magnitude", 
       y = "Impact Value") +
  theme_minimal()

ggplotly(impact_plot)
```

- From the distribution of fatalities associated with tornado magnitude, there is an increasing trend. For lower magnitudes (0-2), fatalities are low and concentrated near zero with very few outliers. Starting at magnitude 3, fatalities show an increase in variability, with some outliers reaching significant numbers. At magnitude 4 and 5, the median fatalities tend to increase with magnitude and there is an outlier over 150 fatalities.

- From the distribution of injuries associated with tornado magnitude, there is an increasing trend. At the lower magnitudes (0-2), injuries are minimal, with medians near 0. Injuries increases significantly from magnitude 3 to magnitude 5. At magnitude 5, there is an outlier indicating extreme cases of over 1,500 injuries. The median at magnitude 5 is lower than the median in the distribution of fatalities with tornado magnitude.

- From the distribution of property loss associated with tornado magnitude, the variability in property loss grows substantially at higher magnitudes, indicates that stronger tornadoes often result in catastrophic property damage. 

- By these three box plots, we can observe that the stronger tornadoes (magnitude 4-5) are associated with exponentially greater impact on fatalities, injuries and property loss. While the lower magnitude tornadoes (magnitude 0-2) are generally less impactful, the significant outliers at higher magnitudes indicates the potential for catastrophic events. These findings highlight the importance of preparedness and mitigation measures for communities in areas prone to high-magnitude tornadoes, as they pose the most severe risks.

Association between tornadoes magnitude with "path_length" and "path_width"

```{r}
impact_plot2 = cleaned %>%
  pivot_longer(cols = c(len, wid), 
               names_to = "metric", 
               values_to = "value") %>%
  ggplot(aes(x = as.factor(mag), y = value, color = metric)) +
  geom_boxplot(outlier.colour = "blue", outlier.size = 0.8) +
  facet_wrap(~metric, 
             scales = "free_y", 
             labeller = labeller(metric = c(
    "len" = "Path Length (miles)",
    "wid" = "Path Width (yards)"
  ))) +
  labs(
    title = "Distribution of Path Length and Width by Tornado Magnitude",
    x = "Tornado Magnitude",
    y = "Value"
  ) +
  theme_minimal()

ggplotly(impact_plot2)
```

- Both path length and path width increase consistently with tornado magnitude, which suggests that stronger tornadoes tend to cover longer distance and have wider destructive zones.

- As magnitude increases, the variability in both path length and width grows significantly. Higher magnitude tornadoes are not only more intense but also less predictable in their extent and size.

- The outliers are more common and extreme at higher magnitudes, which reflects the rare but catastrophic tornadoes that occur at these levels.

Which states have experienced the highest cumulative property losses due to tornadoes?

```{r}
# States with highest average property loss per tornado
state_property_loss <- cleaned |>
  group_by(st) |>
  summarize(avg_loss = mean(loss, na.rm = TRUE)) |>
  arrange(desc(avg_loss)) |>
  slice(1:10) |>
  plot_ly(x = ~st, y = ~avg_loss, type = 'bar') |>
  layout(title = "Top 10 States with Highest Average Property Loss",
         xaxis = list(title = "State"),
         yaxis = list(title = "Average Property Loss (Millions)"))
state_property_loss
```

- **Tennessee (TN)** has the highest average property loss, exceeding **2.5 million dollars**, significantly more than other states.
- **Ohio (OH)** and **North Carolina (NC)** follow with average losses over **1 million dollars**.
- **Louisiana (LA)**, **Arkansas (AR)**, and **Texas (TX)** report moderate losses between **0.5 and 1 million dollars**.
- The variability suggests that factors like infrastructure and tornado intensity contribute to these differences.

What is the relationship between tornado path length, path width, and property loss?

```{r}
# Correlation between path length, width, and property loss
corr <- cleaned |>
  plot_ly(x = ~len, y = ~wid, z = ~loss, color = ~mag) |>
  add_markers() |>
  layout(title = "Correlation Between Path Length, Width, and Property Loss",
         scene = list(xaxis = list(title = "Path Length (miles)"),
                      yaxis = list(title = "Path Width (yards)"),
                      zaxis = list(title = "Property Loss (Millions)")))
corr
```

- Tornadoes with longer paths and wider widths are more likely to result in higher property losses, as shown by the clustering of points in the upper-right corner of the plot.
- Property losses are generally concentrated below **$500 million**, but a few events exceed **$1 billion**.
- Higher magnitude tornadoes (indicated by warmer colors such as yellow) are associated with larger property losses and damages.

What is the total property loss for each state?

```{r}
state_loss_map <- cleaned |>
  group_by(st) |>
  summarize(total_loss = sum(loss, na.rm = TRUE)) |>
  plot_ly(
    type = 'choropleth',
    locations = ~st,
    locationmode = 'USA-states',
    z = ~total_loss,
    colorscale = "Viridis", 
    zmin = 0,               
    zmax = max(cleaned$loss, na.rm = TRUE), 
    colorbar = list(title = "Total Loss (Millions)")
  ) |>
  layout(
    title = "Total Property Loss by State",
    geo = list(scope = "usa")
  )

state_loss_map
```

- **Texas (TX)** reports the highest total property loss, exceeding **$1.5 billion**, due to its frequent tornado activity and large impacted areas.
- **Tennessee (TN)** also exhibits significant losses, reflecting specific high-damage events during the observed period.
- **Southeastern states**, including **Mississippi (MS)** and **Alabama (AL)**, show moderate property losses.
- **Western and Northeastern states** generally experience the lowest total losses, likely due to lower tornado frequencies and intensities in these regions.

